Characteristic polynomials of p-adic matrices

### 2017-01-30 charpoly/charpoly.tex

The main benefit of working with lattices is the following result
\citelist{\cite{caruso-roe-vaccon:14a}*{Lem. 3.4} \cite{caruso-roe-vaccon:15a}*{Prop. 2.2}}.

Let $\chi : M_n(K) \to K_n[X]$ be the characteristic polynomial map.  For a matrix $M \in M_n(K)$
we identify the tangent space $V$ at $M$ with $M_n(K)$ itself, and the tangent space $W$ at $\chi(M)$ with
the space $K_{<n}[X]$ of polynomials of degree less than $n$.  Write $B^-_V(r)$ (resp. $B_V(r)$) for the open (resp. closed)
ball of radius $r$ in $V$.  
Our analysis of the $\chi$ begins with the following result
\citelist{\cite{caruso-roe-vaccon:14a}*{Lem. 3.4} \cite{caruso-roe-vaccon:15a}*{Prop. 2.2}}.

\begin{prop} \label{prop:mainlem}
Suppose that $M \in M_n(K)$ and that $d\chi(M): V \to W$ is surjective.
Let $C$ be a positive real number such that
$B_W(1) \subset d\chi(M)(B_V(C))$.  Then, for all $\rho \in (0, 1]$ and all
$r \in (0, C \cdot \rho^{-1})$, any lattice $H$ such that $B_V^-(\rho r) \subset H \subset B_V(r)$
satisfies:
\begin{equation}
\chi(M + H) = \chi(M) + d\chi(H).
\end{equation}
\end{prop}

\subsection{Computation of the differential}

In light of Proposition \ref{prop:mainlem}, we recall the computation of $d\chi$
\citelist{\cite{caruso-roe-vaccon:14a}*{Appendix B} \cite{caruso-roe-vaccon:15a}*{\S 3.3}}.
Let $\com(M)$ denote the comatrix of $M$; when $M \in \GL_n(K)$ then $\com(M) = \det(M) M^{-1}$.  Then $d\chi_M$ is given
by
\[
d\chi_M: dM \mapsto \tr(\com(X-M) \cdot dM).
\]
%Moreover, $d\chi(M)$ will be surjective exactly when

%\subsection{Properties of the image of $d \chi$}

@@

\begin{prop}
Let $A$ be the subring of $K[X]$ consisting of polynomials $P$ for
which $P(M) \in M_n(\OK)$. Then $d \chi_M \big(M_n(\OK)\big)$ 
(considered as a submodule of $K[X]/\chi_M(X)$) is  stable by 
multiplication by the elements of $A$.
\end{prop}

\begin{proof}
\todo{The image is generated by the entries of $C = \com(X-M)$
and we have $P(M) \cdot C \equiv  P(X) \cdot C \pmod{\chi_M(X)}$.}
\end{proof}

@@

@@

Equality of lattices holds iff it does mod p, so need surjectivity modulo p.  Surjectivity of differential
can be seen: write matrix in Jordan normal form.  Comatrix(X - M) surjective iff when you write it
in Frobenius normal form it has only one block.

>>>>

Let $\chi : M_n(K) \to K_n[X]$ be the characteristic polynomial map.
Our analysis of the precision behavior of $\chi$ rests upon
the computation of its derivative $d\chi$, using \cite{caruso-roe-vaccon:14a}*{Lem. 3.4}.
For a matrix $M \in M_n(K)$ we identify the tangent space $V$
at $M$ with $M_n(K)$ itself, and the tangent space $W$ at $\chi(M)$ with
the space $K_{<n}[X]$ of polynomials of degree less than $n$.
Let $\com(M)$ denote the comatrix of $M$ (when $M \in \GL_n(K)$,
we have $\com(M) = \det(M) M^{-1}$).  Recall
\citelist{\cite{caruso-roe-vaccon:14a}*{Appendix B} \cite{caruso-roe-vaccon:15a}*{\S 3.3}}
that $d\chi(M)$ is given by
\begin{equation} \label{eq:dchi}
d\chi(M): dM \mapsto \tr(\com(X-M) \cdot dM).
\end{equation}
Write $B^-_V(r)$ (resp. $B_V(r)$) for the open (resp. closed) ball of radius $r$ in $V$.

\begin{prop} \label{prop:mainlem}
Suppose that $M \in M_n(K)$, all the Jordan blocks of $M$ have distinct generalized eigenvalues, and
let $C = \max(\lvert \det(M)^{-1} \rvert, 1)$.  \todo{What if $\det(M) = 0$?}. Then, for all $\rho \in (0, 1]$ and all
$r \in (0, C \cdot \rho^{-1})$, any lattice $H$ such that $B_V^-(\rho r) \subset H \subset B_V(r)$
satisfies:
\begin{equation}
\chi(M + H) = \chi(M) + d\chi(H).
\end{equation}
\end{prop}

\begin{proof}
We first show that the distinct generalized eigenvalues condition
guarantees that $d\chi(M): V \to W$ is surjective.  For any $A \in \GL_n(K)$,
the image of $d\chi(M)$ will be the same as the image of $d\chi(AMA^{-1})$ ,
so we may assume that $M$ is in Jordan form.  Letting $dM$ range over
matrices with a single $1$ entry and all other entries $0$, it suffices to show
that the entries of $\com(X-M)$ span $W$.  Suppose that the blocks of $M$
have dimensions $d_1, \dots, d_m$ with corresponding generalized eigenvalues
$\lambda_1, \dots, \lambda_m$.  The only nonzero entries of $\com(X-M)$
are those on or below the diagonal within the Jordan blocks, and within the $i$th
block, each entry along the $j$th subdiagonal is
$\pm(X-\lambda_i)^{d_i - j - 1} \prod_{k \ne i} (X-\lambda_k)^{d_k}$.
As long as the $\lambda_i$ are distinct, these form a basis for $W$.

Recall \cite{caruso-roe-vaccon:15a}*{Def. 3.3} that the \emph{precision polygon}
of $M$ is the lower convex hull of the Newton polygons of the entries of $\com(X-M)$.
Since the precision polygon lies below the Newton polygon \cite{caruso-roe-vaccon:15a}*{Prop. 3.4},
and since the smallest entry of the Newton polygon occurs either with the leading
coefficient or the constant term, $B_W(1) \subset d\chi(M)(B_V(C))$.

Since the coefficients of $\chi(M)$ are given by polynomials in the entries of $M$
with integral coefficients, \cite{caruso-roe-vaccon:15a}*{Prop. 2.2} implies
the conclusion.
\end{proof}

@@

\begin{prop}
Let $A$ be the subring of $K[X]$ consisting of polynomials $P$ for
which $P(M) \in M_n(\OK)$. Then $d \chi_M \big(M_n(\OK)\big)$ 
(considered as a submodule of $K[X]/\chi_M(X)$) is stable by 
multiplication by the elements of $A$.
\end{prop}
\begin{proof}
\todo{The image is generated by the entries of $C = \com(X-M)$
and we have $P(M) \cdot C \equiv  P(X) \cdot C \pmod{\chi_M(X)}$.}
\end{proof}

@@

\begin{prop}
Let $M \in M_n(K)$, $r>0$ and $H \subset K[X] / \chi(X)$ be the image
of $B_V(r)$ under $d\chi(M)$.  Then $X \cdot H = H$.
\end{prop}
\begin{proof}
\todo{Finish proof}
\end{proof}

We can also give a criterion on $M$ that allows an easy description of the image of flat
precision lattices under $d\chi(M)$.
\begin{prop}
Suppose that $M \in M_n(\OK)$, and that the reduction $\bar{M}$ of $M$ modulo $\pi$ has
a cyclic vector.  Then for any $r > 0$, we have $d\chi(M)(B_V(r)) = B_W(r)$.
\end{prop}
\begin{proof}
First note that it suffices to prove the claim for $r=1$ since $d\chi(M)$ is linear.
Next, it suffices to check that $d\chi(M)$ is surjective modulo $\pi$.
Surjectivity of differential
can be seen: write matrix in Jordan normal form.  Comatrix(X - M) surjective iff when you write it
in Frobenius normal form it has only one block.
\end{proof}

@@
